---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
library(arrow)
library(sf)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# geoarrow

<!-- badges: start -->
[![R-CMD-check](https://github.com/paleolimbot/geoarrow/workflows/R-CMD-check/badge.svg)](https://github.com/paleolimbot/geoarrow/actions)
[![Codecov test coverage](https://codecov.io/gh/paleolimbot/geoarrow/branch/master/graph/badge.svg)](https://codecov.io/gh/paleolimbot/geoarrow?branch=master)
<!-- badges: end -->

The goal of geoarrow is to prototype Arrow representations of geometry. This is currently a [first-draft specification](https://github.com/jorisvandenbossche/geo-arrow-spec/blob/geo-arrow-format-initial/format.md) and nothing here should be used for anything except entertainment value.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("paleolimbot/geoarrow")
```

## Write and write to Parquet

This exists for prototyping only, but should work with most things that you throw at it. Notably, sf objects should work out-of-the-box.

```{r}
library(geoarrow)

nc <- sf::read_sf(system.file("shape/nc.shp", package = "sf"))
write_geoarrow_parquet(nc, "nc.parquet")
read_geoarrow_parquet_sf("nc.parquet")
```

You can use any of the schemas to experiment by passing the `schema` argument. These control how the geometry is stored in an Arrow data type (but shouldn't affect the round trip back to R).

```{r}
write_geoarrow_parquet(nc, "nc.parquet", schema = geoarrow_schema_wkb())
read_geoarrow_parquet_sf("nc.parquet")
```

```{r, include = FALSE}
unlink("nc.parquet")
```
