---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# geoarrow

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/geoarrow/geoarrow-r/branch/main/graph/badge.svg)](https://app.codecov.io/gh/geoarrow/geoarrow-r?branch=main)
<!-- badges: end -->

The goal of geoarrow is to leverage the features of the [arrow](https://arrow.apache.org/docs/r/) package and larger [Apache Arrow](https://arrow.apache.org/) ecosystem for geospatial data. The geoarrow package provides an R implementation of the [GeoParquet](https://github.com/opengeospatial/geoparquet) file format of and the draft [geoarrow data specification](https://geoarrow.org), defining extension array types for vector geospatial data.

## Installation

You can install the released version of geoarrow from [CRAN](https://cran.r-project.org/) with:

``` r
install.packages("nanoarrow")
```

You can install the development version of geoarrow from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("geoarrow/geoarrow-r")
```

## Example

The geoarrow package implements conversions to/from various geospatial types (e.g., sf, sfc, s2, wk) with various Arrow representations (e.g., arrow, nanoarrow). The most useful conversions are between the **arrow** and **sf** packages, which in most cases allow sf objects to be passed to **arrow** functions directly after `library(geoarrow)` or `requireNamespace("geoarrow")` has been called.

```{r example}
library(geoarrow)
library(arrow, warn.conflicts = FALSE)
library(sf)

nc <- read_sf(system.file("gpkg/nc.gpkg", package = "sf"))
tf <- tempfile(fileext = ".parquet")

nc |> 
  tibble::as_tibble() |> 
  write_parquet(tf)

open_dataset(tf) |> 
  dplyr::filter(startsWith(NAME, "A")) |>
  dplyr::select(NAME, geom) |> 
  st_as_sf()
```

By default, arrow objects are converted to a neutral wrapper around chunked Arrow memory, which in turn implements conversions to most spatial types:

```{r}
df <- read_parquet(tf)
st_as_sfc(df$geom)
```





