---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
library(arrow)

# hack to view the pretty Arrow C++ printing of a C schema
pretty_type <- function(x) {
  carrow::from_carrow_array(list(schema = x), arrow:::DataType)
}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# geoarrow

<!-- badges: start -->
[![R-CMD-check](https://github.com/paleolimbot/geoarrow/workflows/R-CMD-check/badge.svg)](https://github.com/paleolimbot/geoarrow/actions)
[![Codecov test coverage](https://codecov.io/gh/paleolimbot/geoarrow/branch/master/graph/badge.svg)](https://codecov.io/gh/paleolimbot/geoarrow?branch=master)
<!-- badges: end -->

The goal of geoarrow is to prototype Arrow representations of geometry. This is currently a [first-draft specification](https://github.com/jorisvandenbossche/geo-arrow-spec/blob/geo-arrow-format-initial/format.md) and nothing here should be used for anything except entertainment value. 

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("paleolimbot/geoarrow")
```

## Type examples

```{r}
library(geoarrow)
```

First, specs for the kind of data people are already using:

```{r}
pretty_type(geo_arrow_schema_wkb())
pretty_type(geo_arrow_schema_wkt())
pretty_type(geo_arrow_schema_geojson())
```

...but also Arrow-native forms that don't need to be parsed:

```{r example}
pretty_type(geo_arrow_schema_point())
pretty_type(geo_arrow_schema_point_float32())
pretty_type(geo_arrow_schema_linestring())
pretty_type(geo_arrow_schema_polygon())
pretty_type(geo_arrow_schema_multi(geo_arrow_schema_point()))
pretty_type(geo_arrow_schema_multi(geo_arrow_schema_linestring()))
pretty_type(geo_arrow_schema_multi(geo_arrow_schema_polygon()))
# (I can't get dense or sparse unions to import to the Arrow R package at the moment
# but that's how collections could be stored)
```

This setup gives considerable flexibility as to the coordinate storage format (with a reasonable default whose usage could be mandated for a particular library). For example, s2 cell IDs could be used for points using the same linestring type:

```{r}
pretty_type(
  geo_arrow_schema_linestring(
    point = geo_arrow_schema_point_s2()
  )
)
```

The schemas here use column-level extension types and extension metadata to encode dimension names, CRS information, and a flag to specify that edges should be considered ellipsoidal rather than Cartesian.

```{r}
schema <- geo_arrow_schema_linestring(ellipsoidal = TRUE)
schema$metadata[["ARROW:extension:name"]]
geo_arrow_metadata(schema)
geo_arrow_metadata(schema$children[[1]])
```
