---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# geoarrow

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/geoarrow/geoarrow-r/branch/main/graph/badge.svg)](https://app.codecov.io/gh/geoarrow/geoarrow-r?branch=main)
<!-- badges: end -->

The goal of geoarrow is to leverage the features of the [arrow](https://arrow.apache.org/docs/r/) package and larger [Apache Arrow](https://arrow.apache.org/) ecosystem for geospatial data. The geoarrow package provides an R implementation of the [GeoParquet](https://github.com/opengeospatial/geoparquet) file format of and the draft [geoarrow data specification](https://geoarrow.org), defining extension array types for vector geospatial data.

## Installation

You can install the development version of geoarrow from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("geoarrow/geoarrow-r")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(geoarrow)
```
