---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
library(arrow)
library(dplyr)
library(sf)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# geoarrow

<!-- badges: start -->
[![R-CMD-check](https://github.com/paleolimbot/geoarrow/workflows/R-CMD-check/badge.svg)](https://github.com/paleolimbot/geoarrow/actions)
[![Codecov test coverage](https://codecov.io/gh/paleolimbot/geoarrow/branch/master/graph/badge.svg)](https://codecov.io/gh/paleolimbot/geoarrow?branch=master)
[![R-CMD-check](https://github.com/geoarrow/geoarrow-r/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/geoarrow/geoarrow-r/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of geoarrow is to leverage the features of the [arrow](https://arrow.apache.org/docs/r/) package and larger [Apache Arrow](https://arrow.apache.org/) ecosystem for geospatial data. The geoarrow package provides an R implementation of the [GeoParquet](https://github.com/opengeospatial/geoparquet) file format of and the draft [geoarrow data specification](https://github.com/geopandas/geo-arrow-spec), defining extension array types for vector geospatial data.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("paleolimbot/geoarrow")
```

## Read and Write GeoParquet

Parquet is a compact binary file format that enables fast reading and efficient compression, and its geospatial extension 'GeoParquet' lets us use it to encode geospatial data. You can write geospatial data (e.g., sf objects) to Parquet using `write_geoparquet()` and read them using `read_geoparquet()`.

```{r}
library(geoarrow)

nc <- sf::read_sf(system.file("shape/nc.shp", package = "sf"))
write_geoparquet(nc, "nc.parquet")
read_geoparquet_sf("nc.parquet")
```

You can also use `arrow::open_dataset()` and `geoarrow_collect_sf()` to use the full power of the Arrow compute engine on datasets of one or more files:

```{r}
library(arrow)
library(dplyr)

(query <- open_dataset("nc.parquet") %>%
  filter(grepl("^A", NAME)) %>%
  select(NAME, geometry) )

query %>%
  geoarrow_collect_sf()
```

```{r, include = FALSE}
unlink("nc.parquet")
```
