---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
library(arrow)
library(dplyr)
library(sf)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# geoarrow

<!-- badges: start -->
[![R-CMD-check](https://github.com/paleolimbot/geoarrow/workflows/R-CMD-check/badge.svg)](https://github.com/paleolimbot/geoarrow/actions)
[![Codecov test coverage](https://codecov.io/gh/paleolimbot/geoarrow/branch/master/graph/badge.svg)](https://codecov.io/gh/paleolimbot/geoarrow?branch=master)
<!-- badges: end -->

The goal of geoarrow is to leverage the features of the [arrow](https://arrow.apache.org/docs/r/) package and larger [Apache Arrow](https://arrow.apache.org/) ecosystem for geospatial data. The geoarrow package provides an R implementation of the draft [geoarrow data specification](https://github.com/geopandas/geo-arrow-spec), defining extension array types for vector geospatial data.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("paleolimbot/geoarrow")
```

## Write geospatial data to Parquet

Parquet is a compact binary file format that enables fast reading and efficient compression. You can write geospatial data (e.g., sf objects) to Parquet using `geoarrow_write_parquet()` and read them using `geoarrow_read_parquet()`.

```{r}
library(geoarrow)

nc <- sf::read_sf(system.file("shape/nc.shp", package = "sf"))
write_geoarrow_parquet(nc, "nc.parquet")
read_geoarrow_parquet_sf("nc.parquet")
```

You can also use `arrow::open_dataset()` and `geoarrow_collect_sf()` to use the full power of the Arrow compute engine on datasets of one or more files:

```{r}
library(arrow)
library(dplyr)

(query <- open_dataset("nc.parquet") %>%
  filter(grepl("^A", NAME)) %>%
  select(NAME, geometry) )

query %>% 
  geoarrow_collect_sf()
```

```{r, include = FALSE}
unlink("nc.parquet")
```
