---
title: "C and C++ development guide"
---

```{r, include = FALSE}
library(cpp11)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The geoarrow package exposes a [set of high-level helpers to integrate geospatial data with Arrow Datasets](geoarrow.html) that depends on a low-level C++ to seamlessly and efficiently integrate with other representations of geometry in R and elsewhere. The [format used](format.html) can be easily represented using the [Apache Arrow C Data interface](https://arrow.apache.org/docs/format/CDataInterface.html), exposing vectors of geometries as ABI-stable C structures that can be used without any header other than the two `struct` definitions in [arrow/c/abi.h](https://github.com/apache/arrow/blob/master/cpp/src/arrow/c/abi.h).

We'll use the [narrow](https://paleolimbot.github.io/narrow/) and geoarrow packages to demonstrate the structure of geoarrow arrays and how they can be iterated over in C and C++.

```{r}
library(narrow)
library(geoarrow)
```

We'll also need some test arrays with each of the six types supported in the Arrow-native format.

```{r}
points <- geoarrow_create(
  wk::wkt(c("POINT (30 10)", "POINT (40 30)"))
)

linestrings <- geoarrow_create(
  wk::wkt(
    c(
      "LINESTRING (30 10, 10 30, 40 40)", 
      "LINESTRING (0 0, 10 5)"
    )
  )
)

polygons <- geoarrow_create(
  wk::wkt(
    c(
      "POLYGON ((30 10, 40 40, 20 40, 10 20, 30 10))", 
      "POLYGON (
        (35 10, 45 45, 15 40, 10 20, 35 10),
        (20 30, 35 35, 30 20, 20 30))"
    )
  )
)

multipoints <- geoarrow_create(
  wk::wkt(c("MULTIPOINT (0 1, 2 3)", "MULTIPOINT (0 0, 3 8)"))
)

multilinestrings <- geoarrow_create(
  wk::wkt(
    c(
      "MULTILINESTRING ((30 10, 40 40, 20 40, 10 20, 30 10))", 
      "MULTILINESTRING (
        (35 10, 45 45, 15 40, 10 20, 35 10),
        (20 30, 35 35, 30 20, 20 30))"
    )
  )
)

multipolygons <- geoarrow_create(
  wk::wkt(
    c(
      "MULTIPOLYGON (
        ((30 20, 45 40, 10 40, 30 20)),
        ((15 5, 40 10, 10 20, 5 10, 15 5)))", 
      "MULTIPOLYGON (
        ((40 40, 20 45, 45 30, 40 40)),
        ((20 35, 10 30, 10 10, 30 5, 45 20, 20 35),
          (30 20, 20 15, 20 25, 30 20)))"
    )
  )
)
```

```{r, echo=FALSE}
old_par <- par(mfrow = c(2, 3))
for (array in list(points, linestrings, polygons)) {
  wk::wk_plot(array, col = rgb(0, 0, 0, alpha = 0.5))
}
for (array in list(multipoints, multilinestrings, multipolygons)) {
  wk::wk_plot(array, col = rgb(0, 0, 0, alpha = 0.5))
}

par(old_par)
```

The other piece of setup you may need to follow along is a way to pass the example arrays into C and/or C++ code. If you are writing an R package, you can include the narrow package in your `LinkingTo:` field in your DESCRIPTION; if you are writing an RMarkdown document (like this one), you can set your PKG_CPPFLAGS environment variable to include the "include" directory in the narrow package directory:

```{r}
Sys.setenv(
  PKG_CPPFLAGS = paste0("-I", system.file("include", package = "narrow"))
)
```

This will let you `#include "narrow.h"`, which includes a copy of the Arrow C Data structure definitions and functions to extract them safely from R objects.

```{c, results="hide"}
#include "narrow.h"
```

## Reading geoarrow arrays in R

The easiest way to inspect geoarrow array data is from R using the [wk](https://paleolimbot.github.io/wk/) package. The general structure of an array is available from `wk::wk_vector_meta()`, which is fast (it doesn't iterate over all the points). 

```{r}
wk::wk_vector_meta(points)
```

If you need the geodesic flag or the CRS, you can use `wk::wk_is_geoesic()` or `wk::wk_crs()`:

```{r}
wk::wk_is_geodesic(points)
wk::wk_crs(points)
```

You can access coordinate values as flat arrays using `wk_coords()`:

```{r}
wk::wk_coords(linestrings)
```

The geoarrow package implements `wk::wk_handle()`, so you can use any handler or filter defined in any package. For example, you can apply an affine transform and create a `geos::geos_geometry()` vector in one step (with zero intermediary copies):

```{r}
wk::wk_handle(
  linestrings,
  wk::wk_transform_filter(
    trans = wk::wk_affine_rotate(90),
    handler = geos::geos_geometry_writer()
  )
)
```

You can use the `wk::sfc_writer()` handler to generate sf objects. Note that when using the `wk::wk_handle()` interface, you will have to propagate the CRS and/or geodesic flags yourself. For a guide on how to write filters and handlers in C and C++, see the [programming vignette](https://paleolimbot.github.io/wk/articles/programming.html) from the wk package.

If you need geoarrow/arrow-specific metadata information, you can inspect the schemas directly using the list-like interface of schem aobjects. The most important thing you will have to check is the extension name, which will tell you which bit of compiled code you will need to safely access the array data.

```{r}
points$schema$metadata[["ARROW:extension:name"]]
```

For points, you will also have to check dimensions and the storage type, since depending on the dimensions and/or encoding the underlying arrays could have buffer numbers and/or types.

```{r}
points$schema$format
points$schema$children[[1]]$format
points$schema$children[[1]]$name
```

The above output indicates that the point array is a fixed-size list (size 2) whose child is a float64. The dimension output indicates that the dimension type is xy.

You may also need to access the extension metadata which contains information like the CRS. This metadata is serialized in the raw schema, so you'll need `geoarrow_metadata()` to extract it.

```{r}
geoarrow_metadata(points$schema)
```

## Reading geoarrow arrays using geoarrow.hpp

(TODO...write this!)

## Reading geoarrow arrays in C

If you need to deal with arbitrary vectors of that may have different types or different point encodings, you should `#include "geoarrow.hpp"` and use the helper functions provided there; however, the geoarrow format was designed to be accessible from C code without helpers. If you can make some assumptions about the geometry type and storage type of your input but still want the speed of compiled code, using the C structures directly may be a good fit. As an example, I'm going to print coordinates with information about the containing structures.

### Points

For points encoded as a fixed-size list (the default), you need three pieces of information: the length, the initial offset into the array, and the buffer of `double`s containing the coordinates.

```{c, results="hide"}
#include <R.h>
#include "narrow.h"

SEXP c_point_print(SEXP array_data_xptr) {
  struct ArrowArray* array_data = array_data_from_xptr(array_data_xptr, "array_data");
  
  int coord_size = 2;
  double* coords = (double*) array_data->children[0]->buffers[1];
  coords = coords + (array_data->offset) * coord_size;

  double* coord;
  for (int64_t i = 0; i < array_data->length; i++) {
    coord = coords + (i * coord_size);
    Rprintf("point[%d] (%g %g)\n", i, coord[0], coord[1]);
  }
  
  return R_NilValue;
}
```

Depending on whether or not you have access to a higher-level runtime, you probably want to validate a few things about the schema before accessing elements of the array.

```{r}
point_print <- function(x) {
  x <- as_narrow_array(x)
  
  stopifnot(
    x$schema$format == "+w:2",
    x$schema$children[[1]]$format == "g"
  )
  
  invisible(.Call("c_point_print", x$array_data))
}

point_print(points)
```

For points encoded as a struct, the approach is similar but one buffer per dimension is required.

```{c, results="hide"}
#include <R.h>
#include "narrow.h"

SEXP c_point_struct_print(SEXP array_data_xptr) {
  struct ArrowArray* array_data = array_data_from_xptr(array_data_xptr, "array_data");
  
  double* x = (double*) array_data->children[0]->buffers[1];
  double* y = (double*) array_data->children[1]->buffers[1];
  x = x + array_data->offset;
  y = y + array_data->offset;

  for (int64_t i = 0; i < array_data->length; i++) {
    Rprintf("point[%d] (%g %g)\n", i, x[i], y[i]);
  }
  
  return R_NilValue;
}
```

```{r}
point_struct_print <- function(x) {
  x <- as_narrow_array(x)
  
  stopifnot(
    x$schema$format == "+s",
    x$schema$children[[1]]$format == "g",
    x$schema$children[[2]]$format == "g"
  )

  invisible(.Call("c_point_struct_print", x$array_data))
}

points_struct <- geoarrow_create(
  points,
  schema = geoarrow_schema_point_struct()
)

point_struct_print(points_struct)
```

### Linestrings

In addition to all the information about a point, you also need the offsets into the point array from the outer linestring.

```{c, results="hide"}
#include <R.h>
#include "narrow.h"

SEXP c_linestring_print(SEXP array_data_xptr) {
  struct ArrowArray* array_data = array_data_from_xptr(array_data_xptr, "array_data");
  
  int coord_size = 2;
  double* coords = (double*) array_data->children[0]->children[0]->buffers[1];
  int32_t* coord_offsets = (int32_t*) array_data->buffers[1];
  
  coord_offsets = coord_offsets + array_data->offset;
  coords = coords + coord_offsets[0];

  double* coord;
  for (int64_t i = 0; i < array_data->length; i++) {
    int32_t n_coords = coord_offsets[i + 1] - coord_offsets[i];
    
    for (int32_t j = 0; j < n_coords; j++) {
      coord = coords + coord_size * (coord_offsets[i] + j);
      Rprintf("linestring[%d]->point[%d] (%g %g)\n", i, j, coord[0], coord[1]);
    }
  }
  
  return R_NilValue;
}
```

```{r}
linestring_print <- function(x) {
  x <- as_narrow_array(x)
  
  stopifnot(
    x$schema$format == "+l",
    x$schema$children[[1]]$format == "+w:2",
    x$schema$children[[1]]$children[[1]]$format == "g"
  )
  
  invisible(.Call("c_linestring_print", x$array_data))
}

linestring_print(linestrings)
```

### Polygons

In addition to all the information about a linestring, you also need the offsets into the rings array from the outer polygon.

```{c, results="hide"}
#include <R.h>
#include "narrow.h"

SEXP c_polygon_print(SEXP array_data_xptr) {
  struct ArrowArray* array_data = array_data_from_xptr(array_data_xptr, "array_data");
  
  int coord_size = 2;
  double* coords = (double*) array_data->children[0]->children[0]->children[0]->buffers[1];
  int32_t* coord_offsets = (int32_t*) array_data->children[0]->buffers[1];
  int32_t* ring_offsets = (int32_t*) array_data->buffers[1];
  
  ring_offsets = ring_offsets + array_data->offset;
  coord_offsets = coord_offsets + ring_offsets[0];
  coords = coords + coord_offsets[0];

  double* coord;
  int32_t* coord_offset;
  for (int64_t i = 0; i < array_data->length; i++) {
    int32_t n_seq = ring_offsets[i + 1] - ring_offsets[i];
    coord_offset = coord_offsets + ring_offsets[i];
    
    
    for (int32_t j = 0; j < n_seq; j++) {
      int32_t n_coords = coord_offset[j + 1] - coord_offset[j];
      
      for (int32_t k = 0; k < n_coords; k++) {
        coord = coords + (coord_size * (coord_offset[j] + k));
        Rprintf(
          "polygon[%d]->ring[%d]->point[%d] (%g %g)\n",
          i, j, k, coord[0], coord[1]);
      }
    }
  }
  
  return R_NilValue;
}
```

```{r}
polygon_print <- function(x) {
  x <- as_narrow_array(x)
  
  stopifnot(
    x$schema$format == "+l",
    x$schema$children[[1]]$format == "+l",
    x$schema$children[[1]]$children[[1]]$format == "+w:2",
    x$schema$children[[1]]$children[[1]]$children[[1]]$format == "g"
  )
  
  invisible(.Call("c_polygon_print", x$array_data))
}

polygon_print(polygons)
```

### Multipoints

Iterating over multipoints requires the same code and assertions as for linestrings:

```{r}
linestring_print(multipoints)
```

### Multilinestrings

Iterating over multilinestrings requires the same code and assertions as for polygons:

```{r}
polygon_print(multilinestrings)
```

### Multipolygons

In addition to all the information about a polygon, you also need the offsets into the geometries array from the outer collection. 

```{c, results="hide"}
#include <R.h>
#include "narrow.h"

SEXP c_multipolygon_print(SEXP array_data_xptr) {
  struct ArrowArray* array_data = array_data_from_xptr(array_data_xptr, "array_data");
  
  int coord_size = 2;
  double* coords = (double*) array_data->children[0]->children[0]->children[0]->children[0]->buffers[1];
  int32_t* coord_offsets = (int32_t*) array_data->children[0]->children[0]->buffers[1];
  int32_t* ring_offsets = (int32_t*) array_data->children[0]->buffers[1];
  int32_t* geom_offsets = (int32_t*) array_data->buffers[1];
  
  geom_offsets = geom_offsets + array_data->offset;
  ring_offsets = ring_offsets + geom_offsets[0];
  coord_offsets = coord_offsets + ring_offsets[0];
  coords = coords + coord_offsets[0];

  double* coord;
  int32_t* coord_offset;
  int32_t* ring_offset;
  for (int64_t i = 0; i < array_data->length; i++) {
    int32_t n_geoms = geom_offsets[i + 1] - geom_offsets[i];
    ring_offset = ring_offsets + geom_offsets[i];
    
    for (int32_t j = 0; j < n_geoms; j++) {
      int32_t n_rings = ring_offset[j + 1] - ring_offset[j];
      coord_offset = coord_offsets + ring_offsets[j];
      
      
      for (int32_t k = 0; k < n_rings; k++) {
        int32_t n_coords = coord_offset[k + 1] - coord_offset[k];
        
        for (int32_t l = 0; l < n_coords; l++) {
          coord = coords + (coord_size * (coord_offset[k] + l));
          Rprintf(
            "multipolygon[%d]->polygon[%d]->ring[%d]->point[%d] (%g %g)\n",
            i, j, k, l, coord[0], coord[1]);
        }
      }
    } 
  }
  
  return R_NilValue;
}
```

```{r}
multipolygon_print <- function(x) {
  x <- as_narrow_array(x)
  
  stopifnot(
    multipolygons$schema$format == "+l",
    multipolygons$children[[1]]$schema$format == "+l",
    multipolygons$children[[1]]$schema$children[[1]]$format == "+l",
    multipolygons$children[[1]]$schema$children[[1]]$children[[1]]$format == "+w:2",
    multipolygons$children[[1]]$schema$children[[1]]$children[[1]]$children[[1]]$format == "g"
  )
  
  invisible(.Call("c_multipolygon_print", x$array_data))
}

multipolygon_print(multipolygons)
```

### Metadata

The easiest way to inspect the schema, which encodes the storage type and extra geo-specific extension metadata, is to do it before you drop into C. In particular, `wk::wk_vector_meta()` has been implemented for geoarrow arrays in such a way that it will (1) validate the schema to let you make stronger assumptions about the C structures you are given and (2) deserialize and extract the metadata you might need to reduce the number of problems you have to solve in C.

If you really do need to do this in compiled code, you can access the `struct ArrowSchema` directly. For example, to extract the dimensions from C, you can do this:

```{c, results="hide"}
#include <R.h>
#include "narrow.h"

SEXP c_point_dimensions(SEXP schema_xptr) {
  struct ArrowSchema* schema = schema_from_xptr(schema_xptr, "schema");
  return Rf_mkString(schema->children[0]->name);
}
```

```{r}
.Call("c_point_dimensions", points$schema)
.Call("c_point_dimensions", linestrings$schema$children[[1]])
```

If you need extension metadata from C you will have to iterate over the metadata field AND the extension metadata field. These are both encoded in the same way (by design). This gets verbose quickly, but if you really need to do it you can do so in about 50 lines. The important thing to note is that none of the names or values are null-terminated in the metadata fields.

```{c, results="hide"}
#include <R.h>
#include "narrow.h"

SEXP c_dump_metadata(SEXP schema_xptr) {
  struct ArrowSchema* schema = schema_from_xptr(schema_xptr, "schema");
  
  const char* metadata = schema->metadata;
  
  int64_t pos = 0;
  int32_t n, m, name_len, value_len;
  memcpy(&n, metadata + pos, sizeof(int32_t));
  pos += sizeof(int32_t);

  for (int i = 0; i < n; i++) {
    memcpy(&name_len, metadata + pos, sizeof(int32_t));
    pos += sizeof(int32_t);

    const char* name = metadata + pos;
    pos += name_len;

    memcpy(&value_len, metadata + pos, sizeof(int32_t));
    pos += sizeof(int32_t);

    if (name_len >= 20 && strncmp(name, "ARROW:extension:name", 20) == 0) {
        const char* value = metadata + pos;
        pos += value_len;

        Rprintf("ARROW:extension:name: '%.*s'\n", value_len, value);
        
    } else if (name_len >= 24 && strncmp(name, "ARROW:extension:metadata", 24) == 0) {
      memcpy(&m, metadata + pos, sizeof(int32_t));
      pos += sizeof(int32_t);

      for (int j = 0; j < m; j++) {
        memcpy(&name_len, metadata + pos, sizeof(int32_t));
        pos += sizeof(int32_t);

        const char* ext_name = metadata + pos;
        pos += name_len;

        memcpy(&value_len, metadata + pos, sizeof(int32_t));
        pos += sizeof(int32_t);

        const char* ext_value = metadata + pos;
        pos += value_len;

        if (name_len == 0 || value_len == 0) {
            continue;
        }

        Rprintf(
          "ARROW:extension:metadata/%.*s: '%.*s'\n",
          name_len, ext_name, value_len, ext_value);
      }
    } else {
      pos += value_len;
      continue;
    }
  }
  
  return R_NilValue;
}
```

```{r}
invisible(.Call("c_dump_metadata", points$schema))

with_crs <- geoarrow_create(
  wk::wkt(crs = "EPSG:1234"),
  schema = geoarrow_schema_point()
)
invisible(.Call("c_dump_metadata", with_crs$schema))
```

If you're in R or another higher-level runtime, the preferred approach is to extract the information you need before calling low-level compiled code (or to use the geoarrow.hpp helper).
